The Bot on Speaking Terms: The Effects of Conversation Architecture on Perceptions of Conversational Agents

### 2023-02-17 paper.tex
second pass completed! yay!

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
%Conversational agents employ natural language to interact with human users. While it is broadly accepted that the specifics of speech interactions affect the perception of communicating parties, 
%Previous research has demonstrated a relationship between the structure of conversational agents' natural language interaction and how these agents are perceived by their users. However, 
%there is a lack of systematic understanding of the connections between the perception of agents and elements of conversation architecture. % are not well understood. 
%Furthermore, there is no clear definition of the aspects of perception that are affected and the elements of conversation architecture. In this paper, we present a synthesized framework demonstrating the connections between perceptions of agents and conversation architecture elements, using a taxonomy of eleven perceptions grouped into four categories, and ten elements grouped into three categories. We highlight the need for further research for the perception of agent's ability and perception of social connection with agent's, especially as it relates to the linguistic style of conversational agents.  Also, there is the need to be consistent in the assessment of agents' perceptions, and consider ethical concerns in design.
%[AK Edit] Conversational agents are deliberately designed to interact by simulating natural conversation, and the effectiveness of these interactions is shown to be affected by users' perceptions of an agent. While research on human-human and human-agent communication suggests the association between speech specifics and perceptions of communicating parties, there is a lack of systematic understanding of how an agent’s conversation architecture affects users' perceptions. To address this gap, we present a framework that outlines the relationships between elements of agents’ conversation architecture grouped around dialog strategy, content affectiveness, content style and speech format, and agent characteristics grouped around perceptions of interactions with an agent, agent's ability, sociability, and humanness. Synthesized based on literature review (n=57), the framework demonstrates both the identified relationships and the areas lacking empirical evidence. We discuss the implications of the framework for the conversation design and highlight the inconsistency issues with the domain terminology and measurements.
Conversational agents mimic natural conversation to interact with users. Since the effectiveness of interactions strongly depends on users’ perception of agents, it is crucial to design agents’ behaviors to provide the intended user perceptions. Research on human-agent and human-human communication suggests that speech specifics are associated with perceptions of communicating parties, but there is a lack of systematic understanding of how speech specifics of agents affect users' perceptions. To address this gap, we present a framework outlining the relationships between elements of agents’ conversation architecture (dialog strategy, content affectiveness, content style and speech format) and aspects of users’ perception (interaction, ability, sociability and humanness). Synthesized based on literature reviewed from the domains of HCI, NLP and linguistics (n=57), this framework demonstrates both the identified relationships and the areas lacking empirical evidence. We discuss the implications of the framework for conversation design and highlight the inconsistencies with terminology and measurements.
\end{abstract}

@@

\subsubsection*{Relationship Between Perceptions of Agents and Conversation Architecture}
To study the effect of conversation architecture on users' perceptions (RQ3), we extracted the connections explored in each paper, using the perception aspects and conversation architecture elements developed in the previous data analysis steps. Each connection's effect was also recorded based on whether an association was found in the study and the nature of that association. Out of our review corpus, 265 connections between perceptions of agents and conversation architecture were explored in literature. To analyze the specifics of conversation architecture that affect the perceptions of CAs, 72 connections that did not result in observed relationships were discarded, resulting in 193 relationships for analysis. For example, the connection between the matching style of a CA and user satisfaction was removed because Hoegen et al. \cite{hoegen2019end}\cmt{[31]} did not find a significant difference between the style matching agent and the non-style matching agent for overall interaction satisfaction. Out of the 193 relationships, 10 were relationships based on agents using multiple architecture elements (e.g. \cite{seeger2021chatbots}\cmt{[35]}\cite{volkel2021manipulating}\cmt{[68]}). They were not included in the final framework as we could not attribute the perceptions of agents to the underlying core elements. Overall, 183 relationships between individual architecture elements and perceptions of agents are incorporated into our synthesized framework, visualized as a heatmap to demonstrate their relationships with each other (Figure \ref{fig:heatmap-impact}).

@@

Across the aspects for the perception of interaction, both \textit{usability} and \textit{engagement} were assessed frequently in the reviewed literature, with conversation architecture having effects on these perception aspects across all architecture element categories. The aspect of \textit{satisfaction} was used in some studies to evaluate perception but it is not used as often as engagement or usability. Several papers that assessed user satisfactions were conversation agents within the transaction context, especially within the customer service domain (e.g. \cite{diederich2019emulating}\cmt{[25]}\cite{elsholz2019exploring}\cmt{[61]}\cite{gnewuch2018faster}\cmt{[19]}).

The category of conversation architecture elements with the most identified relationships to perceptions of interaction is \textit{content style}. Specifically for the element of \textit{elaborateness}, users found the use of full sentences more useful than keyword only \cite{haas2022keep}\cmt{[78]}\cite{roy2021users}\cmt{[71]}. Otherwise, the effect of an agent's elaborateness on user's perception of interaction depends on the user's preference \cite{miehle2018exploring}\cmt{[51]}, as well as the topic of discussion \cite{haas2022keep}\cmt{[78]}. As for \textit{formality}, various studies reported significant differences in perceptions of interaction between using casual vs. formal styles. Some users experienced higher engagement with agent interacting with the casual style of the conversation \cite{cox2022does}\cmt{[27]}, while in another study users reported the formal language style as less engaging as it is boring \cite{kim2019comparing}\cmt{[89]}.

@@

There are several conversation architecture elements with little to no explorations to the perception of agent's ability. Specifically, both categories of dialog strategy and content affectiveness are minimally explored in our reviewed corpus, with no explored connections with \textit{response delay} and only one or two connections explored for \textit{social dialog} and \textit{humour}. While the content style category had coverage in some aspects of perceived abilities, it is worth noting that \textit{elaborateness} did not have any explored connections with perceived ability of agents. Further research in these areas is needed to fill these gaps.

%% CW - second pass up to here Feb 15

@@

There are 53 explored connections between conversation architecture elements and the perception of social connection with agents, with majority of them (n=40) resulted in relationships with effects. Interestingly, studies explored more perceptions of social connection with agent related to voice-based CAs (n=34) compared to text-based CAs (n=19), potentially due to the assumption that voice-based agents is more suitable to cultivate human-agent relationships. There are some mixed results for voice-based CAs, as 10 out of the 34 connections did not find any significant results. Half of the neutral results come from the conversation architecture element of \textbf{prosodic cues}. For example, the pitch of agents' voices did not impact the perceived rapport \cite{lubold2016effects}\cmt{[86]} between agent and users or the appropriateness of tone \cite{jestin2022effects}\cmt{[81]}. However, there are also findings that the use of vocal cues made the agent sound more empathetic \cite{tolmeijer2021female}\cmt{[62]}, as well as increased perceived emotional connection \cite{chan2021kinvoices}\cmt{[74]}\cite{kim2020can}\cmt{[24]}. This shows that there are nuanced factors within prosodic cues that impact the perception of social connections that we have not captured in our relationships.

%For humour: Clark et al. \cite{clark2019makes} found that humour is an important conversational characteristic for human-human interactions, but viewed as a novelty feature for human-agent conversations.

Across the themes in this perception category, \textbf{social presence} (n=17) and \textbf{conversation tone} (n=18) have similar number of relationships between architecture elements and perceptions, but \textbf{intimacy} (n=5) has less number of impactful relationships. One reason for this is half of the explored connections did not have any effect on the perception of social connection, such as the use of capitalization \cite{westerman2019believe}\cmt{[9]}, lexical alignment \cite{linnemann2018can}\cmt{[15]}, or social dialogue \cite{lubold2016effects}\cmt{[86]}. For \textbf{social presence}, conversation architecture elements from \textbf{non-linguistic format} category, prosodic cue and text variation, had a number of impactful relationships in the reviewed studies. For example, participants conversing with the agents using kin's voices experienced significantly higher perceived co-presence where there is a closer psychological connection with the agent \cite{chan2021kinvoices}\cmt{[74]}. Also, text-based agents including a delay before responding to users were rated higher in social presence ratings compared to the agent that did not use any response delays \cite{gnewuch2018faster}\cmt{[19]}\cite{gnewuch2022opposing}\cmt{[20]}. Lastly, the use of \textbf{prosodic cues} can impact the perceived \textbf{conversation tone} of the agent as emotionally expressive \cite{zhu2022effects}\cmt{[26]} or persuasive \cite{chan2021kinvoices}\cmt{[74]}, as well as the appropriateness of the tone used by the agent \cite{jestin2022effects}\cmt{[81]}\cite{misu2011toward}\cmt{[83]}.

Looking at the different categories of conversation architecture cues, the heatmap (Figure \ref{fig:heatmap-impact} shows that there are not many relationships in \textbf{linguistic style} that are related to the perceived social connection with CAs. There are not many explored connections in our reviewed corpus between the two, which could be an indication of the assumption that linguistic style does not impact the perception of social connections.
While we discussed the impact of non-linguistic elements of prosodic cues and text variations earlier, the conversation architecture elements in the linguistic content category also had effects on the perception of social connections. Specifically, CAs using of \textbf{affective language} were perceived to be more empathetic \cite{daher2020empathic}\cmt{[58]}\cite{diederich2019emulating}\cmt{[25]}\cite{yang2017perceived}\cmt{[44]} and emotionally expressive \cite{zhu2022effects}\cmt{[26]}, and emotional connection with the agent \cite{lee2019s}\cmt{[55]}\cite{lubis2019positive}\cmt{[43]}.

%add text variation in the discussion

@@

Agent's humanness is the second most explored perception within our reviewed corpus (n=76), with majority of the connections finding impacted relationship between the studied conversation architecture elements and the perception of agent's humanness (n=56). While the connections were explored more for voice-based agent (n=49) as compared to text-based agents (n=27), the actual impacted relationships between the modalities were similar (voice=32, text=24). This is due to larger number of neutral relationships within voice modality, such as the use of affective language for a speech agent did not impact its perceived likeability \cite{hu2022polite}\cmt{[76]}.

Across the themes in this perception category, both \textbf{human-likeness} and \textbf{personality traits} have relationships with almost all conversation architecture elements. Based on the reviewed corpus, it seems like all the conversation architecture elements extracted from literature have some impact on the perception of agent's humanness. However, the impacts can be both positive or negative. In the case of Chan et al's study \cite{chan2021kinvoices}\cmt{[74]}, participated rated the agent using kin voices as significantly more likeable compared to the generic voices, but it was perceived as eerie. As for the perceptions within this category, there are more relationships for perceived personality traits (n=33) vs. human-likeness (n=18). This is because many studies only have one measure for human-likeness (e.g. artificial or human-like), but have multiple aspects of personality to be explored. For example, participant rated the agent using affective language as more polite, as well as reflected that the agent seemed friendly \cite{hu2022polite}\cmt{[76]}.

For the conversation architecture elements, formality in the linguistic style category and prosodic cue in the non-linguistic format category had more effects on the perception of agent's humanness in the reviewed corpus. For \textbf{formality}, agents using casual style is perceived as more human-like with a warm, kind and entertaining personality \cite{cox2022does}\cmt{[27]}\cite{jestin2022effects}\cmt{[81]}\cite{kim2019comparing}\cmt{[89]}. The use of \textbf{prosodic cues} varying pitch, intonation and speech rate of a conversation agent generally had impacts on the perceived humanness instead of personality traits \cite{choi2020nobody}\cmt{[54]}\cite{jestin2022effects}\cmt{[81]}\cite{misu2011toward}\cmt{[83]}. For the architecture elements of \textbf{disfluency}, the type of impact on the perception of humanness depended on the context of the conversation. Studies have found that participants perceived the filler-condition agent as more likeable in the social-oriented situation, but did not find the same effect in task-oriented situations \cite{jeong2019exploring}\cmt{[10]}\cite{wester2015artificial}\cmt{[14]}.

@@

\subsubsection{Inconsistencies in Perception Measures}

A major roadblock in synthesizing the findings from existing academic research is the diversity and inconsistency of measurements towards the anthropomorphized perception of agents. It is difficult to assess whether perception measures should be compared with each other. Also, some measures combined anthropomorphized perceptions into a composite score, which makes it impossible to separate the perceptions into more granular details.

Based on the literature reviewed in this paper, we noticed that similar perception concepts were \textbf{measured in different ways} through different surveys. For example, there are several approaches to measure the perceived \textit{human-likeness} of an agent. A commonly used survey is adapted from the Godspeed questionnaire \cite{bartneck2009measurement}, which measures human-likeness based on user's impression of the agent as fake / natural, machinelike / humanlike, unconscious / conscious, and artificial / lifelike (used by \cite{hoegen2019end}\cmt{[31]}, \cite{jeong2019exploring}\cmt{[10]} and \cite{ouchi2019should}\cmt{[59]}). Another way to measure human-likeness is adapted from Holtgraves et al.'s \cite{holtgraves2007perceiving} questionnaire, which asks users to assess the agent's perceived human-likeness, skillfulness, thoughtfulness, politeness, responsiveness and engagement (used by \cite{diederich2019emulating}\cmt{[25]} and  \cite{gnewuch2018faster}\cmt{[19]}). One study \cite{westerman2019believe}\cmt{[9]} used the Ascend of Man pictorial scale \cite{kteily2015ascent} as the measure of perceived human-likeness. It is unclear whether these different measures of perceived human-likeness are similar enough to be compared with each other.

 In addition to different methods used to measure for the same perception, our detailed analysis revealed that the measures using the same label may have \textbf{different meanings}. In Diedrech et al.'s study \cite{diederich2019emulating}\cmt{[25]}, \textit{empathy} measure is adapted from \cite{yan2013role} assessing whether the CA gives users individual or personal attention. In Daher et al.'s study \cite{daher2020empathic}\cmt{[58]}, they also measure the perception of empathy, but it is using the RoPE Scale \cite{charrier2019rope} with questions like "the robot cares about my feelings" or "the robot comforts me when I am upset." These two different measures of empathy seem to have different underlying meanings, one assessing the personalization aspect of CAs, while the other is assessing the emotional aspect of CAs. Another example is the measurement of \textit{trustworthiness}, with some constructs measuring the level of sensitive information user's are willing to share \cite{dinev2006privacy}, and some measuring the honesty and truthfulness of an agent \cite{lee2017enhancing}.

Lastly, there are \textbf{composite measures} of anthropomorphized perception across different categories, making it difficult to break down perceptions into granular details for analysis. One such example is Ma et al's study \cite{ma2022ask}\cmt{[29]} on different approaches for CAs to reply to users' uncertain queries. UX score is used to measure the perception of an agent, which asks whether the user thinks the CA's response is pleasing / trustworthy / natural / acceptable / shorten the distance between CA and user. While the study has found significant impact of the use of formal language on UX score, it is not possible to breakdown the measure into perceptions of interaction (pleasing, acceptable), social connection (shorten the distance between CA and user), and humanness (natural, trustworthy). Similar for Hoegen et al.'s study \cite{hoegen2019end}\cmt{[31]} on the impact of conversational style matching, it is not possible to separate the overall interaction score into different perceptions as the composite measure contains questions measuring different categories of perceptions like interaction (engaging) and humanness (trust, likable).

In recent years there has been some effort towards unifying the evaluation of conversational agents, such as the work by Finch et al. \cite{finch2020towards} presenting a comprehensive analysis of current evaluation protocols. More work is needed to categorize and consolidate perception measures to make them consistent and comparable across various studies.

\subsubsection{Nuances in Influencing Factors}

While the discussion in the Framework section above focuses on the overall effect of conversational architecture elements on anthropomorphized perceptions,  there are other nuanced influencing factors that impact user perceptions of agents.

Depending on the \textbf{type of conversation} user is having with agents, it may lead to differences in perceptions for similar conversational architecture elements. Cox et al. \cite{cox2022does}\cmt{[27]} found that the perceived competence and appropriateness of tone of different language formality styles depend on the sensitivity of information discussed in the conversation. In this particular study, a formal language style is preferred in medical history discussions as it is matching to the expectation of users, but there is no difference in perception between formal and casual styles when discussing income level and credit scores. In Jeong et al's study \cite{jeong2019exploring}\cmt{[10]}, users found the agent using fillers less intelligent and likable in a task-oriented conversation, but found the same agent using fillers as slight more intelligent and likable in a social-oriented conversation. Another factor that could impact perceptions is the \textbf{anonymity} of conversation. Based on user feedback, Lee et al. \cite{lee2020hear}\cmt{[23]} found that the anonymity of a conversational agent is a driving factor encouraging people to self disclose without having to worried about judgements from the CA. The perception of the agent may be different if the user's identity is not anonymous.

Other than the type of conversation, \textbf{user's characteristics} can also be influencing factors on their perception of agents. For example, the effect of matching user's conversational style on perception of agents depend on user's own conversational style, as participants with high consideration conversational style rated the lexically aligned agent as more trustworthy \cite{hoegen2019end}\cmt{[31]}. In the study by Cox et al. \cite{cox2022does}\cmt{[27]}, participants with lower levels of privacy concerns as well as those who believe in robotic intelligence found the chatbot more enjoyable, warm, competent and appropriate. Similarly, participants with positive attitudes towards computers were significantly more likely to indicate that the usage of fillers by agents was a positive aspect of the conversation, compared to participants with neutral attitudes towards computers \cite{pfeifer2009should}\cmt{[12]}. Lastly, user's prior experience with conversational agents could impact perception of agents. Experienced users perceived the agent with a response delay as lower in social presence compared to the one without delay, as it is seen as inefficient to wait for the CA to respond. However, novice users perceived higher social presence conversing with the CAs using response delays as it is more similar to conversations with human partners \cite{gnewuch2018faster}\cmt{[19]}.

%Designing for personalities, not individual elements

%* The size of the effect for language we observed is clearly higher than the size of the effect for voice. \cite{habler2019effects}\cmt{[63]}

%Combined effects
%* integrated mental model \cite{knijnenburg2016inferring}\cmt{[34]}
%* Anthropomorphized chatbot (text) using affective language (emotional expressions), self-referencing, emoticons, response delays. Resulted in higher perceived anthropomorphism, which in result led to lower negative word of mouth and loss of trust. anthropomorphic design is beneficial not only to initial trust but also to the ongoing trust relationship \cite{seeger2021chatbots}\cmt{[35]}

%For discussions on mental well beings, agent using emojis is rated higher for attitude (emotional connection) . However, for agent with discussing physical well being, agent using emojis is rated as lower for attitude. 
%Attitude has emotional connection, coherent - coherent is an ability perception, vs. emotional connection is a social connection perception. \cite{fadhil2018effect}\cmt{[52]}

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Conversational agents mimic natural conversation to interact with users. Since the effectiveness of interactions strongly depends on users’ perception of agents, it is crucial to design agents’ behaviors to provide the intended user perceptions. Research on human-agent and human-human communication suggests that speech specifics are associated with perceptions of communicating parties, but there is a lack of systematic understanding of how speech specifics of agents affect users' perceptions. To address this gap, we present a framework outlining the relationships between elements of agents’ conversation architecture (dialog strategy, content affectiveness, content style and speech format) and aspects of users’ perception (interaction, ability, sociability and humanness). Synthesized based on literature reviewed from the domains of HCI, NLP and linguistics (n=57), this framework demonstrates both the identified relationships and the areas lacking empirical evidence. We discuss the implications of the framework for conversation design and highlight the inconsistencies with terminology and measurements.
\end{abstract}

@@

\subsubsection*{Relationship Between Perceptions of Agents and Conversation Architecture}
To study the effect of conversation architecture on users' perceptions (RQ3), we extracted the connections explored in each paper, using the perception aspects and conversation architecture elements developed in the previous data analysis steps. Each connection's effect was also recorded based on whether an association was found in the study and the nature of that association. Out of our review corpus, 265 connections between perceptions of agents and conversation architecture were explored in literature. To analyze the specifics of conversation architecture that affect the perceptions of CAs, 72 connections that did not result in observed relationships were discarded, resulting in 193 relationships for analysis. For example, the connection between the matching style of a CA and user satisfaction was removed because Hoegen et al. \cite{hoegen2019end}\cmt{[31]} did not find a significant difference between the style matching agent and the non-style matching agent for overall interaction satisfaction. Out of the 193 relationships, 10 were relationships based on agents modifying multiple architecture elements (e.g. \cite{seeger2021chatbots}\cmt{[35]}\cite{volkel2021manipulating}\cmt{[68]}). They were not included in the final framework as we could not attribute the perceptions of agents to the underlying core elements. Overall, 183 relationships between individual architecture elements and perceptions of agents are incorporated into our synthesized framework, visualized as a heatmap to demonstrate their relationships with each other (Figure \ref{fig:heatmap-impact}).

@@

Across the aspects for the perception of interaction, both usability and engagement were assessed frequently in the reviewed literature, with conversation architecture having effects on these perception aspects across all architecture element categories. The aspect of satisfaction was used in some studies to evaluate perception but it is not used as often as engagement or usability. Several papers that assessed user satisfactions were conversation agents within the transaction context, especially within the customer service domain (e.g. \cite{diederich2019emulating}\cmt{[25]}\cite{elsholz2019exploring}\cmt{[61]}\cite{gnewuch2018faster}\cmt{[19]}).

The category of conversation architecture elements with the most identified relationships to perceptions of interaction is content style. Specifically for the element of \textit{elaborateness}, users found the use of full sentences more useful than keyword only \cite{haas2022keep}\cmt{[78]}\cite{roy2021users}\cmt{[71]}. Otherwise, the effect of an agent's elaborateness on user's perception of interaction depends on the user's preference \cite{miehle2018exploring}\cmt{[51]}, as well as the topic of discussion \cite{haas2022keep}\cmt{[78]}. As for \textit{formality}, various studies reported significant differences in perceptions of interaction between using casual vs. formal styles. Some users experienced higher engagement with agent interacting with the casual style of the conversation \cite{cox2022does}\cmt{[27]}, while in another study users reported the formal language style as less engaging as it is boring \cite{kim2019comparing}\cmt{[89]}.

@@

There are several conversation architecture elements with little to no explorations to the perception of agent's ability. Specifically, both categories of dialog strategy and content affectiveness are minimally explored in our reviewed corpus, with no explored connections with \textit{response delay} and only one or two connections explored for \textit{social dialog} and \textit{humour}. While the content style category had coverage in some aspects of perceived abilities, it is worth noting that \textit{elaborateness} did not have any explored connections with perceived ability of agents. Further research in these areas is needed to fill these gaps in understanding.

@@

The perception of agent's humanness is the second most explored category within our reviewed corpus (n=64), with majority of the connections identified relationship between the studied conversation architecture elements and the anthropomorphized perceptions of agent (n=49). Interestingly, studies explored more perceptions of sociability with agent related to voice-based CAs (n=40) compared to text-based CAs (n=24), potentially due to the assumption that interactions with text-based agents are perceived as less personal and more formal \cite{kocielnik2018designing}. There are some mixed results for voice-based CAs, as 12 out of the 28 connections did not find any significant results. Drilling down to the specific elements, we noticed that the use of \textit{alignment} in voice-based agents did not effect the perception of sociability with agent \cite{healey2013relating}\cmt{[39]}\cite{linnemann2018can}\cmt{[15]}. We didn't have data on text-based agents' perceived social connections with users to compare with voice-based agents, as the studies using alignment and text modality in our corpus did not assess this aspect category of perception.

Across the aspects in this perception category, conversation tone had the most number of identified relationship (n=27) with the perceived social connection with CAs, followed by the aspect of social presence (n=17). There are only 5 relationships found that are linked to the perception aspect of intimacy. One reason for this is half of the explored connections did not have any effect on the perception of sociability with agent, such as the use of capitalization \cite{westerman2019believe}\cmt{[9]}, lexical alignment \cite{linnemann2018can}\cmt{[15]}, or social talk \cite{lubold2016effects}\cmt{[86]}. For the perception aspect of social presence, we noticed a gap in the exploration of conversation architecture elements of formality, alignment, and elaborateness in the content style category, as there are no papers explored at this intersection.

Looking at the different categories of conversation architecture elements, the heatmap (Figure \ref{fig:heatmap-impact} shows that there are not many relationships in content style category that are related to the perceived sociability with CAs in general, with most of the discovered effects concentrated at the intersection of \textit{formality} and conversation tone. Studies have found that CAs using casual style of conversation are perceived as warm, empathetic and friendly \cite{jestin2022effects}\cmt{[81]}\cite{kim2019comparing}\cmt{[89]}, while CAs using formal style of conversation are perceived as polite but lacks empathy \cite{cox2022does}\cmt{[27]}. The rest of the architecture elements and perception aspects of sociability are minimally explored in literature, which indicates the need for more research in this area.

There are a few specific conversation architecture elements that had more identified relationships with perceptions of sociability with agent. Specifically, CAs using of \textit{affective language} were perceived to be more empathetic \cite{daher2020empathic}\cmt{[58]}\cite{diederich2019emulating}\cmt{[25]}\cite{yang2017perceived}\cmt{[44]} and emotionally expressive \cite{zhu2022effects}\cmt{[26]}, as well as being more emotional connected with their users \cite{lee2019s}\cmt{[55]}\cite{lubis2019positive}\cmt{[43]}. Also, different variations of a voice-based CA's \textit{prosody} have effects on the perceived sociability with agent, such as an agent using expressive prosody is assessed as more intimate and more similar with the user \cite{kim2020can}\cmt{[24]}. Lastly, the conversation architecture element of \textit{humour} has identified relationships across the social connection category of perception. Clark et al. \cite{clark2019makes} found that humour is an important conversational characteristic for human-human interactions, but viewed as a novelty feature for human-agent conversations. However, in our review we found that humor has effects on human-agent relationships, as humorous agents are rated as more friendly, intimate, and similar by users compared to non-humorous agents \cite{go2021conversational}\cmt{[80]}\cite{khooshabeh2011does}\cmt{[37]}.

@@

There are 55 explored connections between conversation architecture elements and the perception of sociability with agents, with majority of them (n=38) finding identified relationships in literature. There are significantly more explorations for voice-based agents (n=39) as compared to text-based agents (n=16). Previous studies have found that modality may have an effect on the perception of humanness, as voice-based agents are perceived as more human-like as compared to text-based agents \cite{cho2019effects}. Out of the 39 explored connections for voice-based agents, 14 of them did not result in any relationships. This is especially evident in the architecture element of \textit{affective language} for voice-based agents, as most of the connections resulted in null relationships. For example, when comparing the ratings of human-likeness or likeability for a speech agent employing expressive words to the one not using any, Zhu et al \cite{zhu2022effects}\cmt{[26]} were unable to detect any statistically significant differences in both of the observation study and interaction study.

Across the aspects in this perception category, both human-likeness and personality traits have relationships with almost all the conversation architecture elements. Specifically for the use of \textit{prosody} elements, our review found some opposing effects on the perception of the agent's humanness. In the case of Chan et al's study \cite{chan2021kinvoices}\cmt{[74]}, participated rated the agent using kin voices as significantly more likeable compared to the generic voices, but it was perceived as eerie. This may be a warning indicator to watch out for the uncanny valley effect \cite{mori2012uncanny} and advise CA designers to carefully select conversation architecture elements to elicit anthropomorphized perceptions.

There are a few notable conversation architecture elements related to the perceived humanness of conversational agents. The use of \textit{prosody} such as varying pitch, intonation and speech rate has more identified relationships with the perceived personality traits of an agent as compared to the perceived human-likeness, especially on the aspect of likeability \cite{choi2020nobody}\cmt{[54]}\cite{jestin2022effects}\cmt{[81]}\cite{misu2011toward}\cmt{[83]}. For the architecture elements of \textit{disfluency}, the type of impact on the perception of humanness depended on the context of the conversation. Studies have found that participants perceived the filler-condition agent as more likeable in the social-oriented situation, but did not find the same effect in task-oriented situations \cite{jeong2019exploring}\cmt{[10]}\cite{wester2015artificial}\cmt{[14]}. The use of \textit{alignment} in an agent had a number of identified relationships with the perception of personality traits, with some studies finding CAs that are lexically aligned with a user are more likeable \cite{huiyang2022improving}\cmt{[17]}\cite{linnemann2018can}\cmt{[15]}.

@@

\subsubsection{The Need for Consistent Evaluation Protocols for Perceptions of Agents}

We found a diversity of approaches used to assess perceptions of agents in our reviewed corpus. While we made our best efforts to group perceptions used in literature based on similarity with each other, the inconsistent and composite nature of these perception assessments makes it challenging to synthesize them into a framework that contains homogeneous components in each perception aspect that can be compared with each other.

Based on the literature reviewed in this paper, it is unclear whether perception aspects using similar labels are being evaluated \textit{consistently} through different approaches. For example, we found several methods used to assess the perceived human-likeness of an agent. A commonly used survey is adapted from the Godspeed questionnaire \cite{bartneck2009measurement}, which evaluates human-likeness based on users' impression of the agent as fake / natural, machinelike / humanlike, unconscious / conscious, and artificial / lifelike (used by \cite{hoegen2019end}\cmt{[31]}, \cite{jeong2019exploring}\cmt{[10]} and \cite{ouchi2019should}\cmt{[59]}). Another method used to evaluate human-likeness is adapted from Holtgraves et al.'s \cite{holtgraves2007perceiving} questionnaire, which asks users questions related to the agent's perceived human-likeness, skillfulness, thoughtfulness, politeness, responsiveness and engagement (used by \cite{diederich2019emulating}\cmt{[25]} and  \cite{gnewuch2018faster}\cmt{[19]}). One study \cite{westerman2019believe}\cmt{[9]} used the Ascend of Man scale \cite{kteily2015ascent} with pictures showing the evolution from ape to man, asking users to choose a depiction that best represents the agent's perceived human-likeness. It is unclear whether these different methods are capturing assessments of similar perception aspects that can be compared with each other. Another example is the evaluation of the perception of agent's empathy. In Diedrech et al.'s study \cite{diederich2019emulating}\cmt{[25]}, empathy is assessed by asking users whether the CA is giving users individual or personal attention. In another study that also studies the perception of empathy \cite{daher2020empathic}\cmt{[58]}, the RoPE Scale \cite{charrier2019rope} is used with questions like "the robot cares about my feelings" or "the robot comforts me when I am upset." These two different evaluations of empathy seem to have different underlying meanings, one assessing the personalization aspect of CAs, while the other is assessing the emotional aspect of CAs.

In addition the problem of consistency in evaluating perceptions, there is the issue of  \textit{composite measures} being used in the assessment of users' perceptions. These composite measures collapse multiple aspects of perception into one measurement, making it impossible to break down perceptions into more granular aspects for analysis. One such example is Ma et al's study \cite{ma2022ask}\cmt{[29]} on the different approaches for CAs to reply to users' uncertain queries. A single UX score is used to evaluate the users' perceptions, which composes of questions on whether the user thinks the CA's response is pleasing / trustworthy / natural / acceptable / shorten the distance between CA and user. The UX score encompasses multiple aspects across several perception categories. While this study has found significant effect for the use of formal language on the user rated UX score, it is not possible to understand how the details of formality is related to the perception categories of interaction (pleasing, acceptable), sociability (shorten the distance between CA and user), and humanness (natural, trustworthy). The humanness questionnaire from Holtgraves et al \cite{holtgraves2007perceiving} has a similar problem, evaluating across perception categories of interaction (responsiveness, engagement), ability (skillfulness), sociability (politeness), and humanness (human-likeness, thoughtfulness).

There has been some effort recently towards unifying the evaluation of conversational agents, such as the work by Finch et al. \cite{finch2020towards} presenting a comprehensive analysis of current evaluation protocols. More research is needed in this area to standardize the assessment of perceptions to make them consistent, granular, and comparable across literature.

\subsubsection{Investigate the Relationship Across Perception Aspects of Agents}

There is evidence in literature that perception aspects of agents have effects on each other. For instance, Moussawi et al. \cite{moussawi2021perceptions}\cmt{[36]} conducted a study to understand the correlations between different perceptions related to users' intention to adopt a conversational agent. Specifically, they found a correlation within the category of perceived ability, where an agent's perceived intelligence is positively correlated to the perceived initial trust of the agent. This study also found that users with higher perceived intelligence of an agent are more likely to attribute higher ratings for perceived human-likeness, as well as for the usability and engagement aspects in the perception category of interaction. Lastly, the authors found that perceived humanness have a positive impact on perceived enjoyment, which lead to higher intention to adopt the CA. This study outlines that the perception aspects identified in this paper are not independent of each other.

Correlations between perception aspects are discussed in a few of the papers we reviewed. One of the studies showed that the CA's perceived personality trait of agreeableness has an influence on the users' perception of credibility \cite{andrews2012system}\cmt{[38]}. Seeger et al's study found a similar correlation, where higher perceived anthropomorphism led to lower loss of trust \cite{seeger2021chatbots}\cmt{[35]}. There are also correlations between aspects found within the same perception category, such as a speech agent that is rated higher in perceived human-likeness was associated with higher likability ratings \cite{zhu2022effects}\cmt{[26]}. Also, within the perception category of social connection, one study found that social distance is positively related to the perceived social attraction \cite{westerman2019believe}\cmt{[9]}.

These correlations demonstrate that some perception aspects have effects on each other, either within the same category or across different categories. While there are limited research into this area, further investigations are needed to understand the relationships between various perception aspects.

\subsubsection{Research Directions for the Effects of Conversation Architecture on Users' Perceptions}

Our synthesized framework (Figure \ref{fig:heatmap-impact}) serves as a foundational understanding on the effect of conversation architecture on the perception of agents. It showcases the density of explored relationships between them, as well as highlighting areas that are under-explored. In order to continue building our knowledge, it is important to investigate further into the effect of contextual factors as well as multiple conversation architecture elements on users' perceptions.

There are a few \textit{under-explored areas} in our synthesized framework. The perception category of agent's ability has the least number of explored connections with conversation architecture compared to other perception categories. Specifically, more studies are needed to understand the effects of social dialog, response delay, humour and elaborateness on users' perceptions of agents' ability. For the perception category of sociability with agent, there is a general lack of explored connections with the conversation architecture category of content style, especially for alignment and elaborateness elements. Lastly, more studies on the effect of dialog strategy for conversation architecture elements of initiative and response delay on the perceptions of agent are needed.

Some studies have discussed how differences in \textit{contextual factors} resulted in variations in the perceptions of agents while using similar conversation architecture elements. One of the factors is users' experience with CAs, as Gnewuch et al. \cite{gnewuch2018faster}\cmt{[19]} found that experienced users perceived the agent with a response delay as lower in social presence because it is seen as inefficient to wait for the CA to respond, but novice users perceived higher social presence conversing with the CAs using response delays because it is more similar to conversations with human partners \cite{gnewuch2018faster}\cmt{[19]}. User characteristics is another factor, as the perceived trustworthiness of a style matching agent depended on users' own conversational style \cite{hoegen2019end}\cmt{[31]}. Some other contextual factors that resulted in differences in perceptions of agent include the purpose of conversation (e.g. transactional vs. social) \cite{jeong2019exploring}\cmt{[10]}, anonymity of the conversation \cite{lee2020hear}\cmt{[23]}, and the sensitivity of information discussed in the interaction \cite{cox2022does}\cmt{[27]}. A comprehensive review to identify these contextual factors as well as to understand their effects on the perceptions of agents across various conversation architecture elements would be useful to tailor CA design for specific situations.

As we gain a better understanding about the effects of a single conversation architecture element on the perceptions of agents, we can extend the research to using \textit{multiple conversation architecture elements} in an agent. Several papers in our reviewed corpus incorporated composite elements in an agent, such as the design of an anthropomorphized chatbot using elements such as affective language, emoticons, response delays to assess users' perceptions compared to a non-anthropomorphized chatbot \cite{seeger2021chatbots}\cmt{[35]}. In addition to explore the combined effects of conversation architecture elements on users' perceptions, it would also be interesting to understand the relative importance of each element on perceptions. Some studies analyzed the effect of modifying conversation architecture elements individually, as well as their combined effect on user's perceptions (e.g. \cite{habler2019effects}\cmt{[63]}\cite{lubold2016effects}\cmt{[86]}\cite{zhu2022effects}\cmt{[26]}). Specifically, Habler et al. \cite{habler2019effects}\cmt{[63]} found that the effect for social dialog is higher than the effect for prosody on the perception of agents.
